---
title: "Airway analysis"
output: html_document
date: "2024-11-04"
knit: (function(inputFile, encoding) {
      source('src/helpers.R');
      rmarkdown::render(inputFile,
                        encoding=encoding,
                        output_dir=save_dir)})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory
setwd('../')

# load packages
library(pluralize)

# load data
treatment.groups <- readRDS(file.path(save_dir, "treatmentGroups.rds"))
raw.counts.matrix <- readRDS(file.path(save_dir, "rawCountsMatrix.rds"))
filteredCountsMat <- readRDS(file.path(save_dir, "filteredCountsMat.rds"))
filteredCountsMatUQ <- readRDS(file.path(save_dir, "filteredCountsMatUQ.rds"))
```

## Introduction

This report contains exploratory data analysis results of the 'airways' package RNAseq data.

The data consists of `r ncol(raw.counts.matrix)` samples (`r knitr::combine_words(colnames(raw.counts.matrix))`), and `r length(levels(treatment.groups))` treatment groups (`r knitr::combine_words(levels(treatment.groups), before = "'", sep = " / ", and = "")`).

## Methods

### Filtering, tidying and normalising.

The data were filtered to exclude low-expression counts.
Genes were required to have a minimum of `r paste(min.cpm, pluralize("count", min.cpm))` per million in at least `r min.cpm.fraction` percent of samples. 
After filtering there were `r nrow(filteredCountsMat)` genes retained from the initial `r nrow(raw.counts.matrix)` genes.

In the 'Airway' package data genes are referred to by an accession number (in this case Ensembl ID).
To make it easier to interpret the data, the Ensemble IDs were replaced with the corresponding gene symbol, using the biomaRt package.
In cases where matches were not found, the Ensemble IDs were retained.

Finally, an upper-quartile normalise function was applied to correct for any differences in library depth between samples.

## Results

Upper quartile normalisation increased the consistency of counts between samples (compare figures 1 and 2).

```{r unnormalised, echo=FALSE}
boxplot(log2(1+filteredCountsMat), las=2, cex.axis=0.5, cex=0.5,
        ylab="log2(1+counts)")
```

*Figure 1. Box plot of un-normalised counts by sample*

```{r normalised, echo=FALSE}
boxplot(log2(1+filteredCountsMatUQ), las=2, cex.axis=0.5, cex=0.5,
        ylab="log2(1+counts)")
```

*Figure 2. Box plot of upper-quartile normalised counts by sample*

There was not a strong difference in counts between treatment groups.
While the treatment group had larger variation, there was strong overlap in the distribution of counts between treatments (see figure 3).

```{r count-by-treatment, echo=FALSE}
plot(y = colSums(filteredCountsMat)/10^6, ylab="Counts (millions)", 
     ylim=c(0, max(colSums(filteredCountsMat)/10^6)),
     x = treatment.groups)
```